Subgroup detection in linear growth curve models with generalized linear mixed model (GLMM) trees

### 2024-02-08 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3280 255f4249-2263-4417-b1ea-8498779e27a7
Former-commit-id: eb01daae67197c936d81f48c4903114f9b805270

\abstract{Growth curve models are popular tools for studying the development of a response variable within subjects over time. Heterogeneity between subjects is common in such models, and researchers are typically interested in explaining or predicting this heterogeneity. We show how generalized linear mixed effects model (GLMM) trees can be used to identify subgroups with different trajectories in linear growth curve models. Originally developed for clustered cross-sectional data, GLMM trees are extended here to longitudinal data. The resulting extended GLMM trees are directly applicable to growth curve models as an important special case. In simulated and real-world data, we assess the performance of the extensions and compare against other partitioning methods for growth curve models. Extended GLMM trees perform more accurately than the original algorithm and LongCART, and similarly accurate as structural equation model (SEM) trees. In addition, GLMM trees allow for modeling both discrete and continuous time series, are less sensitive to (mis-)specification of the random-effects structure and are much faster to compute.\\}

@@

<<echo=FALSE>>=
library("glmertree")
load("bag_ids, N = 250, Science.Rda")
load("Science ability data.Rdata")
names(sciedata) <- c("CHILDID", "GENDER", "RACE", "SES", "GMOTOR", "FMOTOR", 
                 "INTERN", "EXTERN", "INTERP", "CONTRO", "FIRKDG", 
                 "AGEBASELINE", "science", "score", "asmtmm", "months")
data <- sciedata[sciedata$CHILDID %in% bag_ids[ , 39], ]
data$CHILDID <- factor(data$CHILDID)
data$GENDER <- factor(data$GENDER)
data$asmtmm <- factor(data$asmtmm)
library("lmerTest")
LMM <- lmer(score ~ months + (1|CHILDID), data = data)
@

@@

\begin{figure}[b]%
\caption{Growth curves of science ability.}
\begin{subfigure}{0.65\textwidth}%
<<fig=TRUE, echo=FALSE, results=hide, fig.width=7, fig.height=7>>=
y <- t(sapply(unique(data$CHILDID), function(x) data$score[data$CHILDID == x]))
x <- t(sapply(unique(data$CHILDID), function(x) data$months[data$CHILDID == x]))
plot(data$months, data$score, type = "n", xlab = "Months after baseline", 
     ylab = "Science ability", xaxt = "n", cex.lab = 1, cex.axis=1,
     xlim = c(0.5, 57^(2/3)), main = "Full sample (N = 250)")
axis(1, at=c(0, 12, 36, 60)^(2/3), labels=c(0, 12, 36, 60), cex.axis = .8)
sapply(1:nrow(x), function(row) lines(x[row, ], y[row, ], col = gray(0.2, alpha = 1/6), lwd = 2))
lines(x = c(min(data$months), max(data$months)), 
     y = c(fixef(LMM)[1], fixef(LMM)[1] + max(data$months)*fixef(LMM)[2]),
     lwd = 3, col = 2)
@
\\\footnotesize{\textit{Note.} Gray lines depict observed individual trajectories, red line depicts average growth curve as estimated with a linear mixed-effect model, comprising a fixed effect of time and a random intercept with respect to individuals. The $x$-axis is not linear in the number of months, because time was scaled as $(\text{\# of months})^\frac{2}{3}$ in order to obtain the approximately linear trajectories shown here. To facilitate interpretation, the labels on the $x$-axes represent values in the original time metric.}
\end{subfigure}%
\label{fig:global_trajectories}
\end{figure}%

@@

Recursive partitioning methods (RPMs), also known as ``trees``, allow for identifying relevant predictors from a potentially (very) large number of covariates. Figure~\ref{fig:LMM_tree_r} shows an example tree, which identified socio-economic status (SES), gross motor skills (GMOTOR) and internalizing problems (INTERN) from a set of 11 socio-demographic and behavioral characteristics of the children, assessed at baseline. Five subgroups were identified, corresponding to the terminal nodes of the tree, each with a different estimate of the fixed intercept and slope, which are presented in Table~\ref{tab:local_coefs}. Groups of children with higher SES also have higher intercepts, indicating higher average science ability. The group of children with lower SES (node~2) is further split based on gross motor skills, with higher motor skills resulting in a higher intercept. The group of children with intermediate levels of SES (node~6) is further split based on internalizing problems, with lower internalizing problems resulting in a higher intercept. The two groups (or nodes) with higher intercepts also have higher slopes, indicating that children with higher baseline ability also gain more ability over time.

\begin{figure}%
\caption{Partitioned growth curves of science ability.}
\begin{subfigure}{1\textwidth}%
<<fig=TRUE, height=9, width=13.5, echo=FALSE>>=
LMM_form <- score ~ months | CHILDID | GENDER + RACE + SES + GMOTOR + FMOTOR + 
  INTERN + EXTERN + INTERP + CONTRO + FIRKDG + AGEBASELINE
testdat <- sciedata[!sciedata$CHILDID %in% bag_ids[ , 39], ]
testdat$CHILDID <- factor(testdat$CHILDID)
testdat$GENDER <- factor(testdat$GENDER)
testdat$asmtmm <- factor(testdat$asmtmm)
## Remove obs with levels of race in test that are not in train
levs <- unique(testdat$RACE) %in% unique(data$RACE)
if (!all(levs)) {
  testdat <- testdat[-which(!testdat$RACE %in% unique(testdat$RACE)[levs]), ]
}

## Fit models
lmm_tree <- lmertree(LMM_form, data = data, maxdepth = 6)
lmm_tree_c <- lmertree(LMM_form, data = data, cluster = CHILDID, maxdepth = 6)
lmm_tree_r <- lmertree(LMM_form, data = data, maxdepth = 6, ranefstart = TRUE)

## Plot tree
plot(lmm_tree_r, which = "growth", ip_args = list(pval = FALSE), 
     nodesize_level = "CHILDID", tp_args = 
       list(xscale = c(0, 60^(2/3)), xaxis.at = c(0, 12, 36, 60)^(2/3),
            xaxis.labs = c(0, 12, 36, 60)), gp = gpar(cex = 1.5))
@
\label{subfig:tree}
\end{subfigure}%
%~
%\begin{subfigure}{0.25\textwidth}%
%\vspace{-2.95cm}
%\hspace{-2.5cm}
%\scriptsize
<<echo=FALSE, results=tex, fig.pos='b'>>=
# library("kableExtra")
# coefs <- coef(lmm_tree_r)
# coefs <- data.frame(cbind(Node = as.numeric(rownames(coefs)), coefs))
# colnames(coefs) <- c("Node", "Intercept", "Slope")
# coefs$Slope <- paste0("$", formatC(coefs$Slope, digits = 3, format = "f"), "$")
# coefs$Intercept <- paste0("$", formatC(coefs$Intercept, digits = 3, format = "f"), "$")
# coefs$Intercept[!grepl("-", coefs$Intercept)] <- paste("$\\:\\:$",
#                                                        coefs$Intercept[!grepl("-", coefs$Intercept)])
# kable(coefs, format = "latex", linesep = "", # linesep command suppressess addlinesep every 5 rows
#         row.names = FALSE, digits = 3L,escape=FALSE, align = c("ccc"), booktabs = TRUE)
@
%\label{subfig:coefs}
%\end{subfigure}
\\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, $y$-axes represent science ability. Gray lines depict observed individual trajectories. Red lines depict the average growth curve within each terminal node, as estimated with a linear mixed-effect model comprising node-specific fixed effects of time and a random intercept with respect to individuals. The $x$-axis is not linear in the number of months, because time was scaled as $(\text{months after baseline})^\frac{2}{3}$ in order to obtain the approximately linear trajectories shown here. To facilitate interpretation, the labels on the $x$-axes represent values in the original time metric.}
\label{fig:LMM_tree_r}
\end{figure}%

@@

The tree in Figure~\ref{fig:LMM_tree_r} has been estimated with generalized linear mixed-effects model trees \citep[GLMM trees; ][]{FokkySmit18}. GLMM trees were originally proposed for subgroup detection in clustered cross-sectional studies, where subjects are nested in treatment centers, classrooms and/or geographical areas, for example. In the current paper we extend GLMM trees, so that they can be applied to partitioning LGCMs. The general idea of GLMM trees is appealing for subgroup detection in almost any type of mixed-effects model. Compared to clustered cross-sectional data, however, longitudinal data may require a different estimation approach: The variance of random effects tends to be higher with longitudinal data, and the predictors of interest tend to be measured at higher levels (e.g., time-invariant covariates). In this paper, we propose and test two extensions of GLMM trees that account for these characteristics. We focus on the specific use case of partitioning LGCMs, but the extensions are critical for a wider range of settings where covariates are measured at higher levels and/or where the random effects have substantial variance.

There are several alternative RPMs that can be used to partition linear growth-curve models (LGCMs): GUIDE \citep{Loh02}, longRPart \citep{AbdoyLeBl02}, GEE-based decision trees \citep{Lee05}, longitudinal interaction trees \cite[IT; ][]{SuyMene11}, SEM trees \citep{BranyOert13}, mixed-effects longitudinal trees \citep[MELT; ][]{EoyCho14} and LongCART \cite{KundyHare19}. Further, the longRPart2 \citep{StegyJaco18} and IT-LT \citep{WeiyLiu20} methods allow for subgroup detection in non-linear growth curve models\footnote{Both IT methods specifically target subgroups with different time-by-treatment interactions, so are not generally applicable for partitioning growth curve models.}.

The main characteristic that sets GLMM trees apart from other methods for partitioning LGCMs is its local-global estimation approach: GLMM trees do not fit a full parametric model in each of the subgroups defined by the terminal nodes of the tree. Instead, fixed-effects parameters are estimated \textit{locally}, using the observations within a terminal node, and the random-effects parameters are estimated \textit{globally}, using all observations. This local-global estimation approach was first proposed by \cite{HajjyBell11} and \cite{SelaySimo12} for trees with constant fits (i.e., intercepts only) in the terminal nodes. With GLMM trees, the approach was generalized to allow for GLMs with any number of parameters in the terminal nodes, thus allowing for non-Gaussian responses and targeted detection of a wide range of possible interaction effects in mixed-effects models \citep{FokkySmit18}.

@@

Given their unbiased variable selection, lower model complexity and computational burden, GLMM trees might be particularly useful for subgroup detection in LGCMs. The next section explains how GLMM trees are estimated and propose adjustments for partitioning longitudinal trajectories. Next, the performance of the proposed adjustments is evaluated: In Study~I, we assess performance in simulated datasets, in Study~II, we compare performance of GLMM trees with that of two other methods: SEM trees and LongCART. In Study~III, we assess performance of the proposed adjustments in existing datasets on children's development of reading, math and science abilities. In the Discussion, we summarize our findings and discuss implications.

@@

Throughout this paper we focus on the case with a continuous, normally-distributed response $y_i$ with constant variance $\sigma_\epsilon^2$. Therefore, the identity function can be used for the link $g$ but generalizations to other response variable types within the GLM are straightforward. In the general notation above, $X_i$ is the $n_i \times (p + 1)$ fixed-effects design matrix for subject i $(i = 1, \dots, N)$, comprising $p$ regressors plus one column of 1s for the intercept. In this paper, we assume that time is the only time-varying covariate of interest (i.e., $p = 1$), but other time-varying covariates can also be included in $X_i$. The number $n_i$ and spacing of observed timepoints may differ between subjects. Ttime-constant covariates are not contained in $X_i$, but enter the model through the subgroup indicator $j$. The value of the fixed-effects parameters $\beta$ (here, intercept and time slope) in GLMM trees are node-specific, their value depending on the subgroup/node $j$ into which subject $i$ falls. Subgroup membership is defined by the values of time-constant covariates\footnote{In principle, it is also possible to use time-varying covariates for partitioning with GLMM trees, but this might not be appropriate for partitioning growth curves, as participants changing subgroups over time yields discontinuous growth curves.}.

As in the standard GLMM \citep[e.g., ][]{PinhyBate00}, GLMM, $Z_i$ is the random-effects design matrix for subject $i$, comprising a subset of columns of Xi, and bi is the corresponding vector of random effects for subject $i$. Finally, $b_i$ is assumed to follow a (possibly multivariate) normal distribution with mean zero and (co)variance $\Sigma$.

The parameters of a traditional GLMM can be estimated, among other techniques, by maximum likelihood (ML) or restricted ML (REML). Thus, when it is known into which node~$j$ each subject~$i$ falls, the GLMM specified by Equation~\ref{eq:GLMMtree} can be fitted ``as usual'', yielding \emph{local} subgroup-specific fixed-effect estimates~$\hat \beta_j$ and \emph{global} random-effect estimates~$\hat b_i$. To infer the subgroup membership for all observations~$i$, the random-effect estimate is treated as a known offset and a GLM tree is estimated using the model-based (MOB) recursive partitioning algorithm of \cite{ZeilyHoth08}. The overall GLMM tree model is then estimated by alternating between estimating the partition (i.e., subgroups or terminal nodes $j$), and estimating the random- and fixed-effects parameters, as per the following algorithm:
%
\begin{enumerate}
\setlength\itemsep{0em}
	\setcounter{enumi}{-1}
	\item Initialize by setting step $r = 0$ and all random-effect estimates $\hat{b}_{i,(r)} = 0$.
	\item Set $r = r+1$. Fit a GLM tree using $Z_{i} \hat{b}_{i,(r-1)}$ as an offset, yielding the partition $j_{(r)}$.
	\item Fit the mixed-effects model $g(\mu_{i}) = X_{i} \beta_{j, (r)} + Z_{i} b_{i, (r)}$ with the partition $j_{(r)}$ from Step~1. Extract the random-effect estimates~$\hat{b}_{i,(r)}$ from the fitted model.
	\item Repeat Steps~1 and~2 until convergence.
\end{enumerate}
%
This initialization simply assumes zero random effects. Convergence of the algorithm is monitored through the log-likelihood of the mixed-effects model fitted in Step~3. Typically, this converges when the partition~$j_{(r)}$ from the GLM tree is the same as $j_{(r-1)}$ from the previous step.

@@

Previous publications on mixed-effects recursive partitioning find that initializing the random-effect estimates with zero yields accurate estimates of subgroup memberships and the final models \citep{HajjyBell11, HajjyBell14, HajjyLaro17, SelaySimo12, FuySimo15, FokkySmit18}. \cite{SelaySimo12} assessed the impact of different initialization values and found only minor differences that decreased with increasing sample size. In \cite{FokkySmit18}, we found initializing estimation of GLMM trees with zero random effects performed well in cross-sectional clustered data. With longitudinal data, however, random effects tend to be more pronounced: Repeated measures on the same subjects tend to be correlated more strongly than observations nested within the same unit in cross-sectional data. If random effects are sizable, the initial assumption of zero random effects could provide an unrealistic starting point that may be difficult to overcome in subsequent iterations. Our expectation is that for partitioning LGCMs, initializing estimation with the random effects instead of the subgroup structure may improve subgroup recovery. Specifically, that means the algorithm starts by estimating the classic version of the mixed-effects model from Equation~\ref{eq:GLMMtree} with just one set of fixed-effects coefficients $\beta$ and all subjects in a single group.

@@

To illustrate, we applied both initialization approaches to the dataset from Figure~\ref{fig:global_trajectories}. In fact, the tree presented in Figure~\ref{fig:LMM_tree_r} was estimated by initializing with the random effects. Initializing estimation assuming zero random effects resulted in the tree in Figure~\ref{fig:LMM_tree}. The split based on gross motor skills in Figure~\ref{fig:LMM_tree_r} was not implemented in  Figure~\ref{fig:LMM_tree}, while additional splits were implemented based on gender, race, internalizing problems and fine motor skills. Considering the relatively large number of subgroups in  Figure~\ref{fig:LMM_tree_r}, some with relatively small sample sizes, this tree may overfit and not generalize well to other samples\footnote{Generalizability will be empirically evaluated using cross validation in Study~III.}.

@@

When partitioning longitudinal data, covariates will often be measured at the subject level (i.e., time-invariant covariates), which should be accounted for in computing the estimated covariance matrix $\hat{J}$. In general, this computation makes use of the scores. By summing the scores within clusters prior to computation of the covariances, so-called \textit{clustered} covariances are obtained, which account for dependence between observations within the same cluster \citep{ZeilyKoll20}. This resembles a GEE-type approach with an independence correlation structure. Our expectation is that in partitioning LGCMs, use of clustered covariances in the parameter stability tests will improve subgroup recovery.

The tree in Figure~\ref{fig:LMM_tree_c} was estimated using cluster-level parameter stability tests. Combining cluster-level parameter stability tests and random-effects initialization yielded the exact same tree structure and parameter estimates. Compared to the trees in Figures~\ref{fig:LMM_tree_r} and \ref{fig:LMM_tree}, the cluster-level parameter stability tests provided the most parsimonious tree structure thus far. Table~\ref{tab:ranef} presents the estimated random-effects parameters for the models fitted thus far. Not surprisingly, the estimated residual standard deviation ($\hat{\sigma}_\epsilon$) differs only very slightly between models. Yet, substantial differences can be seen for the standard deviation of the random intercept ($\hat{\sigma}_\epsilon$): With more subgroups or terminal nodes, less variance is captured by the random effects and more variance is captured by the tree structure.

>>>>

\abstract{Growth curve models are popular tools for studying the development of a response variable within subjects over time. Heterogeneity between subjects is common in such models, and researchers are typically interested in explaining or predicting this heterogeneity. We show how generalized linear mixed-effects model (GLMM) trees can be used to identify subgroups with different trajectories in linear growth curve models. Originally developed for clustered cross-sectional data, GLMM trees are extended here to longitudinal data. The resulting extended GLMM trees are directly applicable to growth curve models as an important special case. In simulated and real-world data, we assess performance of the extensions and compare against other partitioning methods for growth curve models. Extended GLMM trees perform more accurately than the original algorithm and LongCART, and similarly accurate as structural equation model (SEM) trees. In addition, GLMM trees allow for modeling both discrete and continuous time series, are less sensitive to (mis-)specification of the random-effects structure and are much faster to compute.\\}

@@

<<echo=FALSE>>=
library("glmertree")
load("Science ability data.Rdata")
load("sizesScience.Rda")
Ns <- c(250, 1000)
nreps <- 100L
## Get sample previously used
set.seed(42)
for (resp in c("Math", "Reading", "Science")) {
  for (N in Ns) {

    ## Read in data and prepare objects for saving results
    load(paste0(resp, " ability data.Rdata"))
    resp_short <- ifelse(resp == "Math", "math", ifelse(resp == "Reading", "read", "scie"))
    data <- get(paste0(resp_short, "data"))
    data$CHILDID <- factor(data$CHILDID)

    ## Generate samples for each repetition
    bag_ids <- sapply(1:nreps, function(x) sample(unique(data$CHILDID), size = N))
    if (N == 250 && resp == "Science")
      intro_sample <- bag_ids[ , 39]

  }
}
sciedata$CHILDID <- factor(sciedata$CHILDID)
sciedata$GENDER <- factor(sciedata$GENDER)
sciedata$RACE <- factor(sciedata$RACE)
## simplify variable names
names(sciedata) <- gsub("P1", "", gsub("T1", "", gsub("C1", "", names(sciedata))))
names(sciedata)[names(sciedata) == "WKSESL"] <- "SES"
data <- sciedata[sciedata$CHILDID %in% intro_sample, ]
library("lmerTest")
LMM <- lmer(score ~ months + (1|CHILDID), data = data)
@

@@

\begin{figure}[b]%
\caption{Growth curves of science ability.}
\begin{subfigure}{0.65\textwidth}%
<<fig=TRUE, echo=FALSE, results=hide, fig.width=7, fig.height=7>>=
y <- t(sapply(unique(data$CHILDID), function(x) data$score[data$CHILDID == x]))
x <- t(sapply(unique(data$CHILDID), function(x) data$months[data$CHILDID == x]))
plot(data$months, data$score, type = "n", xlab = "Months after baseline", 
     ylab = "Science ability", xaxt = "n", cex.lab = 1, cex.axis=1,
     xlim = c(0.5, 57^(2/3)), main = "Full sample (N = 250)")
axis(1, at=c(0, 12, 36, 60)^(2/3), labels=c(0, 12, 36, 60), cex.axis = .8)
sapply(1:nrow(x), function(row) lines(x[row, ], y[row, ], col = gray(0.2, alpha = 1/6), lwd = 2))
lines(x = c(min(data$months), max(data$months)), 
     y = c(fixef(LMM)[1], fixef(LMM)[1] + max(data$months)*fixef(LMM)[2]),
     lwd = 3, col = 2)
@
\\\footnotesize{\textit{Note.} Gray lines depict observed individual trajectories, red line depicts average growth curve as estimated with a linear mixed-effect model, comprising a fixed effect of time and a random intercept with respect to individuals. The $x$-axis is not linear in the number of months, because time was scaled as $(\text{\# of months})^\frac{2}{3}$ in order to obtain approximately linear trajectories. To facilitate interpretation, the labels on the $x$-axes represent values on the original time scale.}
\end{subfigure}%
\label{fig:global_trajectories}
\end{figure}%

@@

Recursive partitioning methods (RPMs), also known as ``trees``, allow for identifying relevant predictors from a potentially (very) large number of covariates. Figure~\ref{fig:LMM_tree_r} shows an example tree, which identified socio-economic status (SES), gross motor skills (GMOTOR) and internalizing problems (INTERN) from a set of 11 socio-demographic and behavioral characteristics of the children, assessed at baseline. Five subgroups were identified, corresponding to the terminal nodes of the tree, each with a different estimate of the fixed intercept and slope, which are presented in Table~\ref{tab:local_coefs}. Groups of children with higher SES also have higher intercepts, indicating higher average science ability. The group of children with lower SES (node~2) is further split based on gross motor skills, with higher motor skills resulting in a higher intercept. The group of children with intermediate levels of SES (node~6) is further split based on internalizing problems, with lower internalizing problems resulting in a higher intercept. The two groups (or nodes) with higher intercepts also have higher slopes, indicating that children with higher baseline ability may also gain more ability over time.

\begin{figure}%
\caption{Partitioned growth curves of science ability.}
\begin{subfigure}{1\textwidth}%
<<fig=TRUE, height=9, width=13.5, echo=FALSE>>=
LMM_form <- score ~ months | CHILDID | GENDER + RACE + SES + GMOTOR + FMOTOR + 
  INTERN + EXTERN + INTERP + CONTRO + FIRKDG + AGEBASELINE

## Fit models
lmm_tree <- lmertree(LMM_form, data = data)
lmm_tree_c <- lmertree(LMM_form, data = data, cluster = CHILDID)
lmm_tree_r <- lmertree(LMM_form, data = data, ranefstart = TRUE)

## Plot tree
plot(lmm_tree_r, which = "growth", ip_args = list(pval = FALSE), 
     nodesize_level = "CHILDID", tp_args = 
       list(xscale = c(0, 60^(2/3)), xaxis.at = c(0, 12, 36, 60)^(2/3),
            xaxis.labs = c(0, 12, 36, 60)), gp = gpar(cex = 1.5))
@
\label{subfig:tree}
\end{subfigure}%
%~
%\begin{subfigure}{0.25\textwidth}%
%\vspace{-2.95cm}
%\hspace{-2.5cm}
%\scriptsize
<<echo=FALSE, results=tex, fig.pos='b'>>=
# library("kableExtra")
# coefs <- coef(lmm_tree_r)
# coefs <- data.frame(cbind(Node = as.numeric(rownames(coefs)), coefs))
# colnames(coefs) <- c("Node", "Intercept", "Slope")
# coefs$Slope <- paste0("$", formatC(coefs$Slope, digits = 3, format = "f"), "$")
# coefs$Intercept <- paste0("$", formatC(coefs$Intercept, digits = 3, format = "f"), "$")
# coefs$Intercept[!grepl("-", coefs$Intercept)] <- paste("$\\:\\:$",
#                                                        coefs$Intercept[!grepl("-", coefs$Intercept)])
# kable(coefs, format = "latex", linesep = "", # linesep command suppressess addlinesep every 5 rows
#         row.names = FALSE, digits = 3L,escape=FALSE, align = c("ccc"), booktabs = TRUE)
@
%\label{subfig:coefs}
%\end{subfigure}
\\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, $y$-axes represent science ability. Gray lines depict observed individual trajectories. Red lines depict the average growth curve within each terminal node, as estimated with a linear mixed-effect model comprising node-specific fixed effects of time and a random intercept with respect to individuals. The $x$-axis is not linear in the number of months, because time was scaled as $(\text{months after baseline})^\frac{2}{3}$ in order to obtain the approximately linear trajectories shown here. To facilitate interpretation, the labels on the $x$-axes represent values on the original time scale.}
\label{fig:LMM_tree_r}
\end{figure}%

@@

The tree in Figure~\ref{fig:LMM_tree_r} has been estimated with generalized linear mixed-effects model trees \citep[GLMM trees; ][]{FokkySmit18}. GLMM trees were originally proposed for subgroup detection in clustered cross-sectional studies, where subjects are nested in treatment centers, classrooms and/or geographical areas, for example. In the current paper we extend GLMM trees, so that they can be applied to partitioning linear growth-curve models (LGCMs). The general idea of GLMM trees is appealing for subgroup detection in almost any type of mixed-effects model. Compared to clustered cross-sectional data, however, longitudinal data may require a different estimation approach: The variance of random effects tends to be higher with longitudinal data, and the predictors of interest tend to be measured at higher levels (e.g., time-invariant covariates). In this paper, we propose and test two extensions of GLMM trees that account for these characteristics. We focus on the specific use case of partitioning LGCMs, but the extensions are critical for a wider range of settings where covariates are measured at higher levels and/or where the random effects have substantial variance.

There are several alternative RPMs that can be used to partition LGCMs: GUIDE \citep{Loh02}, longRPart \citep{AbdoyLeBl02}, GEE-based decision trees \citep{Lee05}, longitudinal interaction trees \cite[IT; ][]{SuyMene11}, SEM trees \citep{BranyOert13}, mixed-effects longitudinal trees \citep[MELT; ][]{EoyCho14} and LongCART \cite{KundyHare19}. Further, the longRPart2 \citep{StegyJaco18} and IT-LT \citep{WeiyLiu20} methods allow for subgroup detection in non-linear growth curve models\footnote{Both IT methods specifically target subgroups with different time-by-treatment interactions, so are not generally applicable for partitioning growth curve models.}.

The main characteristic that sets GLMM trees apart from other methods for partitioning LGCMs is a local-global estimation approach: GLMM trees do not fit a full parametric model in each of the subgroups defined by the terminal nodes of the tree. Instead, fixed-effects parameters are estimated \textit{locally}, using the observations within a terminal node, while the random-effects parameters are estimated \textit{globally}, using all observations. This local-global estimation approach was first proposed by \cite{HajjyBell11} and \cite{SelaySimo12} for trees with constant fits (i.e., intercepts only) in the terminal nodes. With GLMM trees, the approach was generalized to allow for GLMs with any number of parameters in the terminal nodes, thus allowing for non-Gaussian responses and targeted detection of a wide range of possible interaction effects in mixed-effects models \citep{FokkySmit18}.

@@

Given their unbiased variable selection, lower model complexity and computational burden, GLMM trees might be particularly useful for subgroup detection in LGCMs. The next section explains how GLMM trees are estimated and propose adjustments for partitioning longitudinal trajectories. Next, the performance of the proposed adjustments is evaluated: In Study~I, we assess performance in simulated datasets, in Study~II, we compare performance of GLMM trees with that of two other partitioning methods: SEM trees and LongCART. In Study~III, we assess performance of the proposed adjustments in real datasets on children's development of reading, math and science abilities. In the Discussion, we summarize our findings and discuss implications.

@@

Throughout this paper we focus on the case with a continuous, normally-distributed response $y_i$ with constant variance $\sigma_\epsilon^2$. Therefore, the identity function can be used for the link $g$, but generalizations to other response variable types within the GLM are straightforward. In the general notation above, $X_i$ is the $n_i \times (p + 1)$ fixed-effects design matrix for subject $i$ $(i = 1, \dots, N)$, comprising $p$ time-varying covariates plus one column of 1s for the intercept. In this paper, we assume that time is the only time-varying covariate of interest (i.e., $p = 1$), but other time-varying covariates can also be included in $X_i$. The number $n_i$ and spacing of observed timepoints may differ between subjects. Time-constant covariates are not contained in $X_i$, but enter the model through the subgroup indicator $j$. The value of the fixed-effects parameters $\beta$ (here, intercept and time slope) in GLMM trees are node-specific, their value depending on the subgroup/node $j$ into which subject $i$ falls. In the current paper, subgroup membership is defined by the values of time-constant covariates\footnote{In principle, it is also possible to use time-varying covariates for partitioning with GLMM trees, but this might not be appropriate for partitioning growth curves, as participants changing subgroups over time would yield discontinuous growth curves.}.

As in the standard GLMM \citep[e.g., ][]{PinhyBate00} $Z_i$ is the random-effects design matrix for subject $i$, comprising a subset of columns of $X_i$, and $b_i$ is the corresponding vector of random effects for subject $i$. Finally, $b_i$ is assumed to follow a (possibly multivariate) normal distribution with mean zero and (co)variance $\Sigma$.

The parameters of a traditional GLMM can be estimated, among other techniques, by maximum likelihood (ML) or restricted ML (REML). Thus, when it is known into which node~$j$ each subject~$i$ falls, the GLMM specified by Equation~\ref{eq:GLMMtree} can be fitted ``as usual'', yielding \emph{local} subgroup-specific fixed-effect estimates~$\hat \beta_j$ and \emph{global} random-effect estimates~$\hat b_i$. To infer the subgroup membership for all observations~$i$, the random-effect estimate is treated as a known offset and a GLM tree is estimated using the model-based (MOB) recursive partitioning algorithm of \cite{ZeilyHoth08}. The overall GLMM tree model is then estimated by alternating between estimating the partition (i.e., subgroups or terminal nodes $j$), and estimating the random- and fixed-effects parameters, as per the following algorithm:
%
\begin{enumerate}
\setlength\itemsep{0em}
	\setcounter{enumi}{-1}
	\item Initialize by setting step $r = 0$ and all random-effect estimates $\hat{b}_{i,(r)} = 0$.
	\item Set $r = r+1$. Fit a GLM tree using $Z_{i} \hat{b}_{i,(r-1)}$ as an offset, yielding the partition $j_{(r)}$.
	\item Fit the mixed-effects model $g(\mu_{i}) = X_{i} \beta_{j, (r)} + Z_{i} b_{i, (r)}$ with the partition $j_{(r)}$ from Step~1. Extract the random-effect estimates~$\hat{b}_{i,(r)}$ from the fitted model.
	\item Repeat Steps~1 and~2 until convergence.
\end{enumerate}
%
This initialization simply assumes zero random effects. Convergence of the algorithm is monitored through the log-likelihood of the mixed-effects model fitted in Step~2. Typically, this converges when the partition~$j_{(r)}$ from the GLM tree in Step~1 is the same as $j_{(r-1)}$ from the previous iteration.

@@

Previous studies on mixed-effects recursive partitioning find that initializing the random-effect estimates with zero yields accurate estimates of subgroup memberships and final models \citep{HajjyBell11, HajjyBell14, HajjyLaro17, SelaySimo12, FuySimo15, FokkySmit18}. \cite{SelaySimo12} assessed the impact of different initialization values and found only minor differences that decreased with increasing sample size. In \cite{FokkySmit18}, we found initializing estimation of GLMM trees with zero random effects performed well in cross-sectional clustered data. With longitudinal data, however, random effects tend to be more pronounced: Repeated measures on the same subjects tend to be correlated more strongly than observations nested within the same unit in cross-sectional data. If random effects are sizable, the initial assumption of zero random effects could provide an unrealistic starting point that may be difficult to overcome in subsequent iterations. We expect that for partitioning LGCMs, initializing estimation with the random effects instead of the subgroup structure may improve subgroup recovery. Specifically, that means the GLMM-tree algorithm starts by estimating the classic version of the mixed-effects model from Equation~\ref{eq:GLMMtree} with just one set of fixed-effects coefficients $\beta$ and all subjects in a single group.

@@

To illustrate, we applied both initialization approaches to the dataset from Figure~\ref{fig:global_trajectories}. The tree presented in Figure~\ref{fig:LMM_tree_r} was in fact estimated by initializing with the random effects. Initializing estimation assuming zero random effects resulted in the tree in Figure~\ref{fig:LMM_tree}. The split based on gross motor skills in Figure~\ref{fig:LMM_tree_r} was not implemented in Figure~\ref{fig:LMM_tree}, while additional splits were implemented based on gender, race, internalizing problems and fine motor skills. Table~\ref{tab:ranef} presents the estimated random-effects parameters for the models fitted thus far. Compared to the global LMM in Figure~\ref{fig:global_trajectories}, the trees in Figure~\ref{fig:LMM_tree_r} and \ref{fig:LMM_tree} have very similar residual variances, but clearly lower variance of the random intercept. Part of the inter-individual variation captured by the random effects in the global LMM is thus explained by subgroup-specific fixed effects in the trees.

The pooled standard deviations in Table~\ref{tab:ranef} can also be used to compute effect sizes for the subgroup differences \citep{WestyKenn14}. The effect sizes for Figure~\ref{fig:LMM_tree_r} and Table~\ref{tab:local_coefs} range from $d = \Sexpr{round(min(ds), digits = 2)}$ for nodes~7~and~8, indicating a small to medium negative effect of internalizing symptoms; to $d = \Sexpr{round(max(ds), digits = 2)}$ for nodes~3~and~9, indicating a huge effect of socio-economic status and gross motor skills combined. Yet, these estimates are based on training data only, so can be overly optimistic. In light of the small number of respondents for some of the subgroups in Figure~\ref{fig:LMM_tree_r}, these effects may not generalize well to other samples. Therefore, generalizability of the tree structures obtained will be empirically evaluated using cross validation in Study~III.

@@

When partitioning longitudinal data, covariates will often be measured at the subject level (i.e., time-invariant covariates), which should be accounted for in computing the estimated covariance matrix $\hat{J}$. In general, this computation makes use of the scores. By summing the scores within clusters prior to computation of the covariances, so-called \textit{clustered} covariances are obtained, which account for dependence between observations within the same cluster \citep{ZeilyKoll20}. This resembles a GEE-type approach with an independence correlation structure. We expect that in partitioning LGCMs, use of clustered covariances in the parameter stability tests will improve subgroup recovery.

To illustrate, the tree in Figure~\ref{fig:LMM_tree_c} was estimated using cluster-level parameter stability tests. Compared to the trees in Figures~\ref{fig:LMM_tree_r} and \ref{fig:LMM_tree}, this yielded the most parsimonious tree structure thus far. The random-effects parameters in Table~\ref{tab:ranef} show that this smaller GLMM tree also has slightly higher variance of the random intercept. With less subgroups or terminal nodes, less variance can be captured by the tree structure, and more variance can be captured by the random effects. Fitting a GLMM tree with both cluster-level parameter stability tests and random-effects initialization resulted in the exact same tree structure and parameter estimates. In the next sections, we will evaluate whether cluster-level parameter stability tests and/or random-effects initialization provide more accurate and better generalizable GLMM trees.
